Cross-validation of regression models: mixed model examples (edited v2)

### 2023-10-13 scratch/gm_addition_to_cv-edited_v2.Rmd
tweak code for mixed-model vignette addendum

Data <- data.frame(
  group = factor(rep(1:Nb, each=Nw)),
  x = Ax + rep(1:Nx, length.out = Nw*Nb)
) |>
  within(
    {
      xm  <- ave(x, group, FUN = mean) # within-group mean
      y <- Ay +
        Bb * xm +                  # contextual effect
        Bw * (x - xm) +            # within-group effect
        SDre * rnorm(Nb)[group] +  # random level by group
        SDwithin * rnorm(Nb*Nw)    # random error within groups
    }
  )
```

@@

Data_long <- reshape(Data[1:Nx, ], direction = "long", sep = ".", 
              timevar = "effect", varying = grep("\\.", names(Data[1:Nx, ])))
Data_long$id <- 1:nrow(Data_long)
Data_long <- reshape(Data_long, direction = "long", sep = "_", 
              timevar = "modelcode",  varying = grep("_", names(Data_long)))

Data_long$model <- factor(
  c("~ 1", "~ 1 + x", "~ 1 + x + xm", "~ 1 + I(x - xm)")
  [match(Data_long$modelcode, c("mod0", "mod1", "mod2", "mod3"))]
)
```

```{r plot-fits-mod0}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values from random intercept model."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" &  effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + (1 | group)",
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```
```{r plot-fits-mod1}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values from model with random intercept and x as a predictor."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + x + (1 | group)",
  ylim=c(-15, 35),
  key=list(
    corner=c(0.95, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```
```{r plot-fits-mod2}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values from model with random intercept, x, and the group mean of x as predictors."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + x + xm + (1 | group)",
  ylim=c(4, 16),
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```

```{r plot-fits-mod3}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values from the estimated model generating the data."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + I(x - xm) + (1 | group)",
  ylim=c(4, 16),
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```

>>>>

Data <- data.frame(
  group = factor(rep(1:Nb, each=Nw)),
  x = Ax + rep(1:Nx, length.out = Nw*Nb)
) |>
  within(
    {
      xm  <- ave(x, group, FUN = mean) # within-group mean
      y <- Ay +
        Bb * xm +                  # contextual effect
        Bw * (x - xm) +            # within-group effect
        rnorm(Nb, sd=SDre)[group] +  # random level by group
        rnorm(Nb*Nw, sd=SDwithin)    # random error within groups
    }
  )
```

@@

# prepare for plotting:
Data_long <- reshape(Data[1:Nx, ], direction = "long", sep = ".", 
              timevar = "effect", varying = grep("\\.", names(Data[1:Nx, ])))
Data_long$id <- 1:nrow(Data_long)
Data_long <- reshape(Data_long, direction = "long", sep = "_", 
              timevar = "modelcode",  varying = grep("_", names(Data_long)))
Data_long$model <- factor(
  c("~ 1", "~ 1 + x", "~ 1 + x + xm", "~ 1 + I(x - xm)")
  [match(Data_long$modelcode, c("mod0", "mod1", "mod2", "mod3"))]
)
```

```{r plot-fits-mod0}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values from random intercept model."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" &  effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + (1 | group)",
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```
```{r plot-fits-mod1}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values from model with random intercept and x as a predictor."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + x + (1 | group)",
  ylim=c(-15, 35),
  key=list(
    corner=c(0.95, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```
```{r plot-fits-mod2}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values from model with random intercept, x, and the group mean of x as predictors."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + x + xm + (1 | group)",
  ylim=c(4, 16),
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```
```{r plot-fits-mod3}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values from the estimated model generating the data."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + I(x - xm) + (1 | group)",
  ylim=c(4, 16),
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```
